Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-08 article/article.Rmd
Commit 2024-09-08 22:14

```{r example-day, echo=FALSE, fig.cap="Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modelled threshold for the detection of CE events, which are denoted with red circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\\,\\text{W}/\\text{m}^{2}$", fig.pos='H'}
knitr::include_graphics("../images/P-example-day-5.png")
```

@@

The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

We have to note that the excess irradiation related to the CE events cannot be
directly linked to the total energy balance of the atmosphere. The net solar
radiation of the region is not increased, but is rather redistributed through the CE
events.  This is also depicted by the ECE irradiance values, which exceed the
equivalent cloud-free irradiance by a significant amount.

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-1} }
        }
    \caption{Time series for the period 1994 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrences normalized with the highest median
value, which occurs in June, depicting a clear seasonal cycle.  Although CE events
are present throughout the year, the most active months are May and June. During the
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence, the
seasonally varying relative position of the sun, as well as the local landscape
characteristics that may influence the formation of the clouds.  Unfortunately, lack
of detailed data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

>>>>

```{r example-day, echo=FALSE, fig.cap="Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modelled threshold for the detection of CE events, which are denoted with red circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\\,\\text{W}/\\text{m}^{2}$.", fig.pos='H'}
knitr::include_graphics("../images/P-example-day-5.png")
```

@@

We must note that the excess irradiation from the CE events that is received at the
surface does not affect the balance of the energy budget in the atmosphere. The
incoming solar radiation in the region is not increased by the OI but is rather
redistributed through the CE events.  This is also depicted by the ECE irradiance
values, which exceed the equivalent cloud-free irradiance by a significant amount.

\begin{figure}% [h!]
        {\centering
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-2} }\\
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-3} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-1} }
        }
    \caption{Time series of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation at Thessaloniki for the period 1994 – 2023. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the number of occurrences of the CE events in percent, calculated
relative to the available one-minute data in each month.  Although CE events are
present throughout the year, showing a clear seasonal cycle, the most active months
are May and June, with their median representing about $3\,\%$ of the data.  The
least frequent events (about $1\,\%$) occur in August and during winter (December --
February).  This seasonality is a combined effect of different factors, among them
the types of clouds, their frequency of occurrence, the seasonally varying relative
position of the sun, as well as the local landscape characteristics that may
influence the formation of the clouds.  For example, during May and June the frequent
formation of cumulus clouds in the area leads to a significant increase in the CE
events.  However, the lack of detailed data in the area on cloud characteristics does
not allow further analysis. Finally, the interannual variability of the monthly CE
events is quite high as manifested by the size of the boxes and the large monthly
extremes, especially in the summer. Investigation of the causes of this variability
is beyond the scope of this study. It could possibly be related to the cloud cover
variability in winter and autumn over southern Europe, associated with the North
Atlantic Oscillation (NAO)
circulation
